---
title: "College Scorecard Cross-Sectional Analysis"
author: "Zachary Tipton & Miles Richard"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: pdf
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 200
)
library(dplyr)
library(fixest)
library(ggplot2)
library(readr)
library(scales)
library(stringr)
library(tidyr)
```

```{r}
# Load and clean data
df <- read_csv("data/college_scorecard/college_scorecard.csv", show_col_types = FALSE) |>
  mutate(
    share_low_income = as.numeric(share_low_income),
    mean_earnings_10yr_after = as.numeric(mean_earnings_10yr_after),
    online_only = if_else(online_only %in% c(1, "1", "Online only"), 1, 0),
    ownership = factor(ownership),
    state = factor(state),
    high_completion = completion_rate >= 0.60
  )
```

## Introduction

This report uses the College Scorecard dataset from the U.S. Department of Education to study how college characteristics relate to student outcomes.
I focus on whether colleges with higher completion rates tend to have higher median earnings ten years after entry.
This is an important question because prospective students may care about both finishing school and the labor market returns associated with attending different institutions.

## 1. Familiarizing the Reader with the Data

I begin by describing the distribution of a few key variables in the dataset.
The main outcome I study is `median_earnings_10yr_after`.
I also summarize `completion_rate` and a binary indicator for whether a school has a completion rate of at least 60 percent.

```{r}
earnings_mean <- mean(df$median_earnings_10yr_after, na.rm = TRUE)
earnings_median <- median(df$median_earnings_10yr_after, na.rm = TRUE)
completion_mean <- mean(df$completion_rate, na.rm = TRUE)
share_high_completion <- mean(df$high_completion, na.rm = TRUE)
```

The average median earnings ten years after entry are `r dollar(earnings_mean)`, and the median is `r dollar(earnings_median)`.
The average completion rate is `r percent(completion_mean, accuracy = 0.1)`.
About `r percent(share_high_completion, accuracy = 0.1)` of colleges in the sample have a completion rate of at least 60 percent.

```{r}
#| fig.cap: "Distribution of median earnings 10 years after entry"
#| label: fig-earnings-hist
ggplot(df, aes(x = median_earnings_10yr_after)) +
  geom_histogram(bins = 30, fill = "#2C6E49", color = "white") +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    x = "Median earnings 10 years after entry",
    y = "Number of colleges",
    title = "Median earnings vary substantially across colleges"
  ) +
  theme_light(base_size = 14)
```

```{r}
#| fig.cap: "Distribution of college completion rates"
#| label: fig-completion-hist
ggplot(df, aes(x = completion_rate)) +
  geom_histogram(bins = 30, fill = "#7A9E7E", color = "white") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Completion rate",
    y = "Number of colleges",
    title = "Completion rates differ widely across institutions"
  ) +
  theme_light(base_size = 14)
```

## 2. Investigating the Relationship Between Two Variables

My main bivariate question is whether colleges with higher completion rates also have higher median earnings ten years after entry.
One possible motivation is that schools with stronger academic support, better resources, or more prepared students may both graduate more students and produce better labor market outcomes.

```{r}
plot_df <- df |>
  select(completion_rate, median_earnings_10yr_after) |>
  drop_na()

corr_completion_earnings <- cor(
  plot_df$completion_rate,
  plot_df$median_earnings_10yr_after
)

est_bivariate <- feols(
  median_earnings_10yr_after ~ completion_rate,
  data = df,
  vcov = "HC1"
)

effect_10pp <- coef(est_bivariate)["completion_rate"] * 0.10
```

The correlation between completion rate and median earnings is `r round(corr_completion_earnings, 3)`.
The bivariate regression suggests that a 10 percentage point increase in completion rate is associated with about `r dollar(effect_10pp)` higher median earnings, on average.

```{r}
#| fig.cap: "Completion rates and post-college earnings"
#| label: fig-completion-earnings
ggplot(plot_df, aes(x = completion_rate, y = median_earnings_10yr_after)) +
  geom_point(alpha = 0.45, color = "#355070") +
  geom_smooth(method = "lm", se = TRUE, color = "#BC4749") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    x = "Completion rate",
    y = "Median earnings 10 years after entry",
    title = "Schools with higher completion rates tend to have higher earnings"
  ) +
  theme_light(base_size = 14)
```

## 3. Adding Additional Controls

The bivariate relationship may partly reflect omitted variables.
For example, colleges serving more low-income students may face different labor market and completion outcomes, and more selective or more expensive schools may also differ in ways that affect earnings.
To address this concern, I add controls for student economic background, institution type, online status, admissions selectivity, and school size.

```{r}
est_controls <- feols(
  median_earnings_10yr_after ~ completion_rate + share_low_income +
    admission_rate + log(n_students) + online_only + ownership,
  data = df,
  vcov = "HC1"
)

coef_compare <- tibble(
  model = c("Bivariate", "With controls"),
  effect_of_10pp = c(
    coef(est_bivariate)["completion_rate"] * 0.10,
    coef(est_controls)["completion_rate"] * 0.10
  )
)
```

After adding controls, the estimated effect of a 10 percentage point increase in completion rate is:

```{r}
coef_compare
```

This comparison shows whether the simple bivariate relationship was overstating or understating the association between completion and later earnings.
In the final version of the writeup, interpret how much the coefficient changes and whether it remains statistically meaningful.

## 4. Interaction Effects

Next, I test whether the relationship between completion rates and earnings differs by college ownership.
This interaction is useful because the returns associated with completion may vary across public, private nonprofit, and private for-profit institutions.

```{r}
est_interaction <- feols(
  median_earnings_10yr_after ~ completion_rate * ownership +
    share_low_income + admission_rate + log(n_students) + online_only,
  data = df,
  vcov = "HC1"
)
```

The interaction model allows the slope on completion rate to differ across ownership types.
In the final writeup, describe whether the completion-earnings relationship appears stronger for some sectors than for others.

## 5. Forecast Model

Finally, I split the data into training and testing samples to evaluate out-of-sample prediction.
The goal is not only to explain earnings, but also to assess how well a model forecasts unseen colleges.

```{r}
set.seed(5753)

forecast_df <- df |>
  select(
    median_earnings_10yr_after, completion_rate, share_low_income,
    admission_rate, n_students, online_only, ownership
  ) |>
  drop_na()

n <- nrow(forecast_df)
train_index <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- forecast_df[train_index, ]
test_df <- forecast_df[-train_index, ]

forecast_model <- feols(
  median_earnings_10yr_after ~ completion_rate + share_low_income +
    admission_rate + log(n_students) + online_only + ownership,
  data = train_df
)

test_df <- test_df |>
  mutate(
    prediction = predict(forecast_model, newdata = test_df),
    prediction_error = median_earnings_10yr_after - prediction,
    squared_error = prediction_error^2
  )

mspe <- mean(test_df$squared_error)
mae <- mean(abs(test_df$prediction_error))
```

The test-sample mean squared prediction error is `r comma(round(mspe, 0))`.
The test-sample mean absolute prediction error is `r dollar(mae)`.

```{r}
#| fig.cap: "Prediction errors by share of low-income students"
#| label: fig-prediction-errors
ggplot(test_df, aes(x = share_low_income, y = prediction_error)) +
  geom_point(alpha = 0.5, color = "#3A5A40") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#BC4749") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    x = "Share of low-income students",
    y = "Prediction error",
    title = "Forecast errors may vary across school characteristics"
  ) +
  theme_light(base_size = 14)
```

In the final report, explain whether the forecast model performs reasonably well and whether prediction errors are systematically larger for schools serving particular student populations.

## Conclusion

This project studies how institutional characteristics are related to long-run earnings outcomes in the College Scorecard data.
The main analysis focuses on the relationship between completion rates and median earnings, then examines how that relationship changes after adding controls and interactions.
The forecasting exercise provides an additional way to evaluate whether the model captures meaningful variation in college outcomes.
