---
title: "Regression in Practice"
author: "Zachary Tipton"
format: pdf
---
```{r setup}
library(ggplot2) # install.packages("ggplot2")
library(fixest)  # install.packages("fixest")
library(arrow)   # install.packages("arrow")
```

## Intro

This R day is going to have you apply different regression strategies presented in topic 04. 
An emphasis will be placed on interpreting and presenting regression results. 
For this assignment, I have assembled a dataset consisting of homes in Massachusetts. 

There are a lot of variables in this dataset, so please see the separete codebook for details.

Because this data is large (almost 2 million rows), I am using the `parquet` file format. 
You can load this with the `read_parquet` function from the `arrow` package. 
You might have to change your working directory in RStudio to the folder where this `.qmd` file is located.

```{r}
# setwd("04-Regression_in_Practice/R-day/")
df <- read_parquet("data/MA_parcels_sample.parquet")
head(df)
```

## Analysis 
### Basic exploration of variables

One of the main variables we will use in this analysis is the assessed value of the home (`total_value`) and the size of the land (`lot_size_acres`).
Like good data analysts, let's create histograms of both variables. 
Comment on the distribution. 

```{r}
ggplot(df, aes(x = total_value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(
    title = "Histogram of Assessed Home Value",
    x = "Total Value",
    y = "Count"
  ) +
  theme_minimal()

ggplot(df, aes(x = lot_size_acres)) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "white") +
  labs(
    title = "Histogram of Lot Size",
    x = "Lot Size (Acres)",
    y = "Count"
  ) +
  theme_minimal()

```

*Answer:*
Both distributions are right-skewed, with a long tail of high values. 


Now, let's look at a few explanatory variables. 
Use the `table` function to explore the distribution of `n_rooms` and make a histogram of `z_mls_acres`. 

```{r}
table(df$n_rooms)

ggplot(df, aes(x = z_mls_acres)) +
  geom_histogram(bins = 50, fill = "orange", color = "white") +
  labs(
    title = "Histogram of Minimum Lot Size Zoning",
    x = "z_mls_acres",
    y = "Count"
  ) +
  theme_minimal()
```

*Answer:*
Also right-skewed, with a long tail of high values, with a large spike at 0.8 acres.

### Relationship between lot size and building value

We are going to explore the relationship between the lot size and the value of the building. 
We will make a scatter plot to start our analysis.
However, there is way too many points to plot in the dataset, so we will sample a random 10000 rows.

```{r}
sample_df <- df[sample(1:nrow(df), 25000), ]
plot(sample_df$lot_size_acres, sample_df$building_value, 
     xlab = "Lot Size (Acres)", ylab = "Building Value", 
     main = "Scatter Plot of Lot Size vs Building Value")
```

#### Bivariate regression

First, regress building value on the lot size linearly using `feols` from the `fixest` package. 
Interpret the regression coefficient and comment on it's statistical significance.
Make sure to use `vcov = "HC1"` to get robust standard errors.

```{r}
biv_mod <- feols(building_value ~ lot_size_acres, data = sample_df, vcov = "HC1")
summary(biv_mod)
```

*Answer:*
A 1 acre increase in lot size is associated with a $117,408.3 +/- $3260.29 increase in building value. The result is statistically significant at the 1% level.



#### Forecasting

Now, let's try and show what our forecasted values will be. 
To do so, let's create a `data.frame` containing a grid of values of `lot_size_acres`. 
We can do that with `seq(0, 4, 0.01)`.
Call this `pred_grid`.
Create forecasts from your simple linear model using `predict(..., newdata = pred_grid)`.
Store the forecasts back into the `pred_grid` data frame.

To create a plot, first copy your `plot` call above to create the sample scatterplot.
Then, use the `lines` function to overlay your forecasted line of best fit. 
Give this line a distinct color and make the line thicker by adding the arguments `col = "red", lwd = 2`.

```{r}
pred_grid <- data.frame(
  lot_size_acres = seq(0, 4, 0.01)
)

pred_grid$pred_building_value <- predict(biv_mod, newdata = pred_grid)

plot(
  sample_df$lot_size_acres,
  sample_df$building_value,
  pch = 16,
  cex = 0.4,
  col = rgb(0, 0, 0, 0.15),
  xlab = "Lot Size (Acres)",
  ylab = "Building Value",
  main = "Building Value vs. Lot Size with Linear Fit"
)

lines(
  pred_grid$lot_size_acres,
  pred_grid$pred_building_value,
  col = "red",
  lwd = 2
)
```


#### Forecasting with confidence intervals

Now, use your regression to predict the assessed value for a home on a 1.5 acre lot. 
Add the option `se.fit = TRUE` to your `predict` call to get a standard error for the predicted value. 
Present your result with a 95\% confidence interval.

```{r}
pred_15 <- predict(
  biv_mod,
  newdata = data.frame(lot_size_acres = 1.5),
  se.fit = TRUE
)

predicted_value <- pred_15$fit
se_pred <- pred_15$se.fit

ci_lower <- predicted_value - 1.96 * se_pred
ci_upper <- predicted_value + 1.96 * se_pred

predicted_value
ci_lower
ci_upper
```

*Answer:* 
Our predicted value for a home on a 1.5 acre lot is $391,049.10, with a 95% confidence interval of [$383,392.30, $398,706.00].

#### Polynomial

Estimate a new model with a 4th-order polynomial of `lot_size_acres`. 
You can do this simply with the `poly` function.
Forecast this model using `pred_grid` and create a new forecast plot similar to above. 
Add lines for both the linear fit (in red) and the polynomial fit (in blue)

```{r}
poly_mod <- feols(building_value ~ poly(lot_size_acres, 4), data = df, vcov = "HC1")
summary(poly_mod)

pred_grid$pred_poly <- predict(poly_mod, newdata = pred_grid)

plot(
  sample_df$lot_size_acres,
  sample_df$building_value,
  pch = 16,
  cex = 0.4,
  col = rgb(0, 0, 0, 0.15),
  xlab = "Lot Size (Acres)",
  ylab = "Building Value",
  main = "Building Value vs. Lot Size"
)

lines(
  pred_grid$lot_size_acres,
  pred_grid$pred_building_value,
  col = "red",
  lwd = 2
)

lines(
  pred_grid$lot_size_acres,
  pred_grid$pred_poly,
  col = "blue",
  lwd = 2
)
```

#### Adding other predictors

Now, let's try to add the number of rooms and the building area to our polynomial model.
Let's create indicatro variables for each value of `n_rooms` using the `i` function.
Control for building_area linearly.

```{r}
rooms_poly_mod <- feols(
  building_value ~ poly(lot_size_acres, 4) + i(n_rooms) + building_area,
  data = df,
  vcov = "HC1"
)

summary(rooms_poly_mod)
```

Answer the following questions:

1. What is the omitted category for `n_rooms`? 
    
    n_rooms::3 is the omitted category, meaning that the coefficients on the other `n_rooms` indicators are interpreted relative to homes with 3 rooms.

2. Interpret in words the coefficient on `n_rooms::7` and comment on it's statistical significance.
    
    The coefficient on `n_rooms::7` is $57,785.60, meaning that homes with 7 rooms are estimated to be worth $57,785.60 more than homes with 3 rooms, holding lot size and building area constant (ceteris paribus). This result is statistically significant at the 1% level.

3. We could have run a regression controlling linearly for the number of rooms (treating it as a continuous measure). Do your estimates for `i(n_rooms)` make you comfortable with making that simplifying assumption?

    No, the relationship between the number of rooms and building value does not appear to be linear. The coefficients on `i(n_rooms)` do not increase by a constant amount as the number of rooms increases, which suggests that treating `n_rooms` as a continuous variable would not capture the true relationship well.

4. What is the estimated return on square footage for homes in Massachusetts? To help put this result in context, mulitply this coefficient by  the standard deviation of `building_area`. Your interpretation will be a for "1 standard deviation increase in `building_area`"  
    
```{r}
sd(df$building_area, na.rm = TRUE)
coef(rooms_poly_mod)["building_area"]
coef(rooms_poly_mod)["building_area"] * sd(df$building_area, na.rm = TRUE)
```

*Answer:*

A one standard deviation increase in building area is associated with about a $73,471 increase in building value, holding lot size and room indicators fixed.



#### Fixed Effects

One concern we might have with interpreting our regression results is that the number of rooms a house has is likely related to what town the home is in. 
Let's add a set of town indicators to limit our regression to compare homes "in the same town", so to speak.
It turns out there are 98 different towns in our sample, so this will end up slowing down our regression quite a bit. 
Instead, we will specify the `town` variable after a vertical bar, `|`, in our formula. 
It will look like `y ~ x | town`. 
`feols` will estimate these effects very quickly, but note their coefficients are not displayed in the summary. 

Modify the previous regression model to add town "fixed effects" (i.e. a set of indicators for each town).

```{r}
rooms_poly_fe_mod <- feols(
  building_value ~ poly(lot_size_acres, 4) + i(n_rooms) + building_area | town,
  data = df,
  vcov = "HC1"
)

summary(rooms_poly_fe_mod)

```

How does this result change our estimate on `building_area`?

```{r}
coef(rooms_poly_mod)["building_area"]
coef(rooms_poly_fe_mod)["building_area"]
```

*Answer:*

The estimate for `building_area` is slightly different when we include town fixed effects. The coefficient is now $83.69, which is higher than the previous estimate of $57.11. 


### Zoning Regulation

Zoning regulations limit the kind of development that can occur on a parcel.
We have a bunch of zoning measures including:
- `z_mls_acres`, the minimum lot size (the required amount of land needed to build a house).
- `z_permit_2fam`, whether 2 family homes are allowed to be built "By Right", "By Permit", or "Not Allowed".
- `z_residential`/`z_nonresidential`/`z_mixed` that says what kind of development can be built on a parcel: residential, commercial, or both. 

Using regression methods, try to present a case that more stringent regulations (higher minimum lot size) is associated with larger observed lot sizes. 
You should write up your results in a paragraph and include a figure. 
Make sure to interpret the statistical significance of your results.

*Answer:* 

```{r}
zoning_mod <- feols(
  lot_size_acres ~ z_mls_acres + z_mls_acres^2 + z_mls_acres^3,
  data = df,
  vcov = "HC1"
)

summary(zoning_mod)

zoning_grid <- data.frame(
  z_mls_acres = seq(0, 2, 0.05)
)

zoning_grid$pred_lot_size <- predict(zoning_mod, newdata = zoning_grid)

plot(
  pred_lot_size ~ z_mls_acres,
  data = zoning_grid,
  type = "l",
  col = "red",
  lwd = 2,
  xlab = "Minimum Lot Size Regulation",
  ylab = "Predicted Lot Size"
)
```

The regression shows a positive and statistically significant relationship between stricter minimum lot size regulation and larger observed lot sizes. However, the predicted lot size is always below the minimum lot size regulation.